---
title: "Class 5: Data Vis with ggplot"
author: "Joseph (PID:A16788229)"
format: pdf
---

## Intro to ggplot

There are many graphics system in R (was to make figures and plots). These include "base" R plots.Today we will focus mostly on **ggplot2** package.

Let's start with a plot of a simple in-built dataset called `cars`.

```{r}
cars
```
```{r}
plot (`cars`)
```


Let's see how we can make this figure using **ggplot**. First I need to install this package on my computer. To install any R package I use the function 'install.packages()'. 

> I will run 'instal.packages("ggplot2") in my R console not this quarto document!

Before I can use any functions add on packages I need to load the package from my "library()" with the `library(ggplot2)` call. 

```{r}
library(ggplot2)
ggplot(cars) 
```


All ggplot figures have at least 3 things (called layers). These include: 

- **data** (the inplut dataset I wat to plot from)
- **aes** (the aesthetic mapping of the data to my plot)
- **geoms** (the geom_point(), geom_line() etc. that I want to draw).

```{r}
ggplot(cars) +
  aes (x=speed, y=dist) +
  geom_point()
```

Let's add a line to show the relationship here: 

```{r}
ggplot(cars) +
  aes (x=speed, y=dist) +
  geom_point() + 
  geom_smooth(method=lm) + 
  theme_bw() +
  labs(title="My First GGPLOT")

```

>Q1 Which geometric layer should be used to create scatter plots? 

geom_point()

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)

```



> How many genes are in the data set? 

```{r}
nrow(genes)
```

A first plot of data set 

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State)+ 
  geom_point() +theme_bw()+
  scale_colour_manual( values=c("yellow","pink","red")) + labs(title="Gene Expression Changes Upon Drug Treatment ", x= "Control Expression", y="Drug Presence Expression", caption="BIMM 143", subtitle= "Just another scatter plot from ggplot2") 
  
  
```

> Q: Use the table () function on the state column of this data.frame to find out how many 'up' regulated genes there are. What is your answer? 

```{r}
table(genes$State)
```
> Q:Using your values above and 2 significant figures. What fraction of total genes is up-regulated in this dataset?

```{r}
round(table(genes$State)/nrow(genes),4)
```

```{r}
n.tot <- nrow(genes)
vals<- table(genes$State)

vals.percent <- vals/n.tot * 100
round (vals.percent, 2)
```
> Q1. For which phases is data visualization important in our scientific workflows? 
All of the above

>Q2. True or False? The ggplot2 package comes already installed with R? 
FALSE

>Q3. Which plot types are typically NOT used to compare distributions of numeric variables? 
Network graphs

>Q4. Which statement about data visualization with ggplot2 is incorrect? 
ggplot2 is the only way to create plots in R

> Q5. Which geometric layer should be used to create scatter plots in ggplot2?
geom_point()

>Q6. Use the nrow() function to find out how many genes are in this dataset. What is your answer? 
5196

>Q7. Use the colnames() function and the ncol() function on the genes data frame to find out what the column names are (we will need these later) and how many columns there are. How many columns did you find? 
4

>Q8. Use the table() function on the State column of this data.frame to find out how many ‘up’ regulated genes there are. What is your answer? 
127

>Q9. Using your values above and 2 significant figures. What fraction of total genes is up-regulated in this dataset? 
2.44


**Combining Plots**
```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```
```{r}
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)
```

```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.8)+ 
  scale_size_area(max_size = 10) + labs(title = "GDP VS Life Expectancy", x= "Life Expectancy", y= "GDP Per Capita")
```



```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)

```


```{r}
ggplot(gapminder_1957) + 
  aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop) +
  geom_point(alpha=0.7) + 
  scale_size_area(max_size = 10) + labs(title = "GDP VS Life Expectancy", x= "Life Expectancy", y= "GDP Per Capita")
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007) 

ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.8) + 
  scale_size_area(max_size = 10) +labs(title = "GDP VS Life Expectancy", x= "Life Expectancy", y= "GDP Per Capita")+
  facet_wrap(~year)
```
**BAR Charts**
```{r}
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5
```

```{r}
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop, fill=lifeExp))+labs(title="Populations of Countries", x=" Country", y=" Population" )
```


